\bf An HMM-based part-of-speech tagger for Turkish

### 2015-12-09 paper/paper.tex
Write on further research

The project consists of two executables: (1) \texttt{create-model} that reads
the data from the corpus and parses it into a form from which we can easily
construct an \texttt{HMM}. Unfortunately, it is highly impractical---if not
impossible---to directly serialize the \texttt{HMM} datatype since it has three
fields containing functions \footnote{See
  http://stackoverflow.com/questions/17785916/can-haskell-functions
  -be-serialized for further information.}. We consider this a fault in the
design of the \texttt{Data.HMM} package and urge further development; this has
almost no effect on performance, but it obfuscates the code.

@@

I should note that it is not possible to give a complete account of the
program's logic in this chapter although we explain the most salient components.
The source code is in the \texttt{ayberkt-final_project} repository and it is
well-commented. Should the reader have any concerns regarding implemenetation
they could find it convenient to consult the source code.

@@

This function is called several times with lists of values we want to
\emph{count} (such as the list of tag bigrams, or the list of all tagged words)
et cetera. Pre-computing the counts is almost mandatory since frequencies (i.e.,
counts) will need to be accessed many times by the viterbi algorithm when
finding the most probable path through the model. Once the necessary maps are
generated, they are written to files in the directory \texttt{model}.

@@

The executable \texttt{run-tag} is where the actual tagging happens. The
frequencies we have pre-computed in \texttt{create-model} are passed in to the
probability function which we give below.
\begin{lstlisting}[
  language=Haskell,
  mathescape,
  columns=fullflexible,
  basicstyle=\small\ttfamily
]
probability :: (Ord a, Ord b) $\Rightarrow$ (a, b) $\rightarrow$ Map a Int $\rightarrow$ Map (a, b) ℤ $\rightarrow \mathscr{L}$
probability (x, y) c$_1$ c$_2$ = if xCount == 0 || yCount == 0
                                   then logFloat $\epsilon$
                                   else (logFloat yCount) / (logFloat xCount)
  where yCount = fromIntegral (findWithDefault 0 (x, y) c₂) ∷ Double
        xCount = fromIntegral (findWithDefault 0 x c₁) ∷ Double
\end{lstlisting}
This is a straightforward implementation of probability relying on a given count
map. One thing to note is that it returns $\epsilon$ probability (i.e., the
smallest value a \texttt{LogFloat}\footnote{A float in the log domain.} can have) if a given element is not found.

@@

When we are creating the functions required by \texttt{HMM} we are storing the
map returned by a call to \texttt{freqMap} in the lexical closure of
\texttt{probability} by making use Haskell's currying so that we can eloquently
define the two crucial functions as follows:
\begin{lstlisting}[
  language=Haskell,
  mathescape,
  columns=fullflexible,
  basicstyle=\small\ttfamily
]
    transFn s$_1$ s$_2$ = probability (s$_1$, s$_2$) tagFreqs tagBigramFreqs
    outFn s e = probability (e, s) wordFreqs taggedWordFreqs
\end{lstlisting}
The probability functions used to create \texttt{transFn} and \texttt{outFn}
hold the count maps \texttt{tagBigramFreqs} and \texttt{taggedWordFreqs} in
their lexical closure, respectively.

Finally we declare a new \texttt{HMM} as follows:
\newpage
\begin{lstlisting}[
  language=Haskell,
  mathescape,
  columns=fullflexible,
  basicstyle=\small\ttfamily
]
      newHMM = HMM { states      = possibleTags ∷ [POS]
                     , events      = ws ∷ [String]
                     , initProbs   = initProbFn
                     , transMatrix = transFn
                     , outMatrix   = outFn
                     }
\end{lstlisting}

This is a simple record\footnote{A record is a datatype declaration in Haskell
  similar to a struct in C.} containing the probability functions we
defined.\footnote{For some reason, the author of the \texttt{Data.HMM} library
  named the fields (i.e., attribute), that are of function types, using names
  that end with \texttt{-Matrix} although they are \emph{not} matrices. The
  reader should not be confused by this.}

@@

We have shown that it is possible---though not ideal---to build a POS tagger for
Turkish---a canonical agglutinative language---that is based entirely on a
bigram HMM. In this section I analyze the accuracy of the tagger and discuss
simple improvements I can make, as well as its fundamental limitations fixing
which would require a significant rewrite of the tagger. Let us first look at
the accuracy of this model and propose possible fixes that do not fundamentally
change the structure of the model.

@@

We have tested the POS tagger with 6075 words from a small portion of the corpus
that I reserved for testing (i.e., did not use for tranining the HMM). The POS
tagger tagged 4514 of these words correctly (74.3 \%). Here is a brief list of
known problems:

@@

  \item[Duplications:] The Turkish language contains many duplications.
  Syntactically speaking, a duplication behaves like a word itself. In the
  treebank, duplications are denoted with two halves joint the with an
  underscore. Since there is not an easy way to detect whether two sequential
  words are reduplications or not, we are giving duplications in an incorrect
  format to the HMM.  We predict that (based on informal tests with the model)
  that correction of this problem would significantly increase the accuracy.

  \item[Abundance of independent clauses:] It is common in the Turkish language
  to have multiple independent clauses in a given sentence. It is possible that
  this decreases the accuracy of the HMM: instead of a tag $t$ being followed by
  a tag $t'$, we get a lot of $t, <clause>, t'$ which obscures the word order we
  are interested in. Since we may always assign the \textsc{Punc} tag to a comma
  with full certainty, we may simply ignore commas as far as probability is
  concerned and tag them with \textsc{Punc}.

@@

In English we entertain the notion of \emph{syntactic correctness}, which is
more or less a formalism of the correct ordering of words. A speaker of English
that has synactic competence in the language \citep{bachman1990} may know that a
sentence is well-formed, relying on his knowledge of syntactic rules. However,
for the \textit{words}, knowledge of morphological rules is not adequate to
conclude that a word is made up of correctly ordered morphemes; they have to
rely on their lexical knowledge to verify that the morphological form they
derived/inflected is a valid one. Consider the word ``satisfiableness'' which is
not a valid word according to several dictionaries, although being derived with
valid suffixes. The correct form of this concept is ``satisfiability''.

In Turkish, the concept of word validness is a lot like syntactic correctness;
the speaker may rely their knowledge of morphological rules to \emph{generate}
new words. In fact Turkish has \textbf{recursive} morphology \citep{kabak2001}
so a word can be inflected/derived in into infinitely many different forms. This
is why computational processing of Turkish can never fully escape its
morphology.

@@

A salient possibility for further research is the incorporation of word endings
into the model using regular expressions. It is trivial to express Turkish
suffixes, that change their form as not to upset the major vowel harmony in a
given word. We may say with probably more than $90\%$ probability that a given
word has a certain POS looking whether its ending satisfies regular expressions.

\bibliographystyle{apalike}
\bibliography{your_bibliography}

>>>>

The project consists of two executables: (1) \texttt{create-model} that reads
the data from the corpus and parses it into a form from which one can easily
construct an \texttt{HMM}. Unfortunately, it is highly impractical---if not
impossible---to directly serialize the \texttt{HMM} datatype since it has three
fields containing functions \footnote{See
  http://stackoverflow.com/questions/17785916/can-haskell-functions
  -be-serialized for further information.}. I consider this a fault in the
design of the \texttt{Data.HMM} package and urge further development; this has
almost no effect on performance, but it obfuscates the code.

@@

I should note that it is not possible to give a complete account of the
program's logic in this chapter although I explain the most salient components.
The source code is in the \texttt{ayberkt-final\_project} repository and it is
well-commented. Should the reader have any concerns regarding implemenetation
they could find it convenient to consult the source code.

@@

This function is called several times with lists of values designate to be
\emph{counted} (such as the list of tag bigrams, or the list of all tagged words)
et cetera. Pre-computing the counts is almost mandatory since frequencies (i.e.,
counts) will need to be accessed many times by the viterbi algorithm when
finding the most probable path through the model. Once the necessary maps are
generated, they are written to files in the directory \texttt{model}.

@@

The executable \texttt{run-tag} is where the actual tagging happens. The
pre-computed frequencies  in \texttt{create-model} are passed in to the
probability function which I give below.
\begin{lstlisting}[
  language=Haskell,
  mathescape,
  columns=fullflexible,
  basicstyle=\small\ttfamily
]
probability :: (Ord a, Ord b) $\Rightarrow$ (a, b) $\rightarrow$ Map a Int $\rightarrow$ Map (a, b) ℤ $\rightarrow \mathscr{L}$
probability (x, y) c$_1$ c$_2$ = if xCount == 0 || yCount == 0
                                   then logFloat $\epsilon$
                                   else (logFloat yCount) / (logFloat xCount)
  where yCount = fromIntegral (findWithDefault 0 (x, y) c₂) ∷ Double
        xCount = fromIntegral (findWithDefault 0 x c₁) ∷ Double
\end{lstlisting}
This is a straightforward implementation of probability relying on a given count
map. One thing to note is that it returns $\epsilon$ probability (i.e., the
smallest value a \texttt{LogFloat}\footnote{A float in the log domain.} can have) if a given element is not found.

@@

When I am creating the functions required by \texttt{HMM} I am storing the
map returned by a call to \texttt{freqMap} in the lexical closure of
\texttt{probability} by making use Haskell's currying so that the functions can
be eloquently defined as follows:
\begin{lstlisting}[
  language=Haskell,
  mathescape,
  columns=fullflexible,
  basicstyle=\small\ttfamily
]
    transFn s$_1$ s$_2$ = probability (s$_1$, s$_2$) tagFreqs tagBigramFreqs
    outFn s e = probability (e, s) wordFreqs taggedWordFreqs
\end{lstlisting}
The probability functions used to create \texttt{transFn} and \texttt{outFn}
hold the count maps \texttt{tagBigramFreqs} and \texttt{taggedWordFreqs} in
their lexical closure, respectively.

Finally a new \texttt{HMM} is declared as follows:
\begin{lstlisting}[
  language=Haskell,
  mathescape,
  columns=fullflexible,
  basicstyle=\small\ttfamily
]
      newHMM = HMM { states      = possibleTags ∷ [POS]
                     , events      = ws ∷ [String]
                     , initProbs   = initProbFn
                     , transMatrix = transFn
                     , outMatrix   = outFn
                     }
\end{lstlisting}

This is a simple record\footnote{A record is a datatype declaration in Haskell
  similar to a struct in C.} containing the probability functions I have
defined.\footnote{For some reason, the author of the \texttt{Data.HMM} library
  named the fields (i.e., attribute), that are of function types, using names
  that end with \texttt{-Matrix} although they are \emph{not} matrices. The
  reader should not be confused by this.}

@@

I have shown that it is possible---though not ideal---to build a POS tagger for
Turkish---a canonical agglutinative language---that is based entirely on a
bigram HMM. In this section I analyze the accuracy of the tagger and discuss
simple improvements I can make, as well as its fundamental limitations fixing
which would require a significant rewrite of the tagger. Let us first look at
the accuracy of this model and propose possible fixes that do not fundamentally
change the structure of the model.

@@

I have tested the POS tagger with 6075 words from a small portion of the corpus
that I reserved for testing (i.e., did not use for tranining the HMM). The POS
tagger tagged 4514 of these words correctly (74.3 \%). Here is a brief list of
known problems:

@@

  \item[Duplications:] The Turkish language contains many duplications.
  Syntactically speaking, a duplication behaves like a word itself. In the
  treebank, duplications are denoted with two halves joint the with an
  underscore. Since there is not an easy way to detect whether two sequential
  words are reduplications or not, I am giving duplications in an incorrect
  format to the HMM.  I predict that (based on informal tests with the model)
  that correction of this problem would significantly increase the accuracy.

  \item[Abundance of independent clauses:] It is common in the Turkish language
  to have multiple independent clauses in a given sentence. It is possible that
  this decreases the accuracy of the HMM: instead of a tag $t$ being followed by
  a tag $t'$, there are a lot of $t, <clause>, t'$ which obscures the word order
  we are interested in. Since it is possible to always assign the \textsc{Punc}
  tag to a comma with full certainty, I may simply ignore commas as far as
  probability is concerned and tag them with \textsc{Punc}.

@@

In English, there exists a notion of \emph{syntactic correctness}, which is more
or less a formalism of the correct ordering of words. A speaker of English that
has synactic competence in the language \citep{bachman1990} may know that a
sentence is well-formed, relying on his knowledge of syntactic rules. However,
for the \textit{words}, knowledge of morphological rules is not adequate to
conclude that a word is made up of correctly ordered morphemes; they have to
rely on their lexical knowledge to verify that the morphological form they
derived/inflected is a valid one. Consider the word ``satisfiableness'' which is
not a valid word according to several dictionaries, although being derived with
valid suffixes. The correct form of this concept is ``satisfiability''.

In Turkish, the concept of word validness is a lot like syntactic correctness;
the speaker may rely their knowledge of morphological rules to \emph{generate}
new words. In fact Turkish has \textbf{recursive} morphology \citep{kabak2001}
so a word can be inflected/derived in into infinitely many different forms. This
is why computational processing of Turkish can never fully escape its
morphology. In the case of the described tagger, the \texttt{outProbFn} function
becomes increasingly unreliable as the morphological complexity of a word
increases; for words with more than six-seven suffixes, it is unlikely that the
word has been seen before.

@@

A salient possibility for further research is the incorporation of word endings
into the model using regular expressions. It is trivial to express Turkish
suffixes, that change their form as not to upset the major vowel harmony in a
given word. It almost \emph{certain} that a given word has a certain POS looking
whether its ending satisfies regular expressions.

As a more paradigmatically challenging remark, I suggest the possibility of
tackling Turkish from a Distributed Morphology \citep{halle1993distributed}
approach. With a theory that views syntax and morphology as instances of the
same generative engine, it might be possible to expose to true nature of
agglutinative languages hence dramatically increasing the accuracy and reducing
the complexity such a program. \bibliographystyle{apalike}
\bibliography{your_bibliography}
